Need Fulfillment During Intergroup Contact: Three Experience Sampling Studies

### 2023-05-31 Methods-and-Results.tex
typos

After receiving ethical approval from the University Masked for Peer
Review, we recruited 23 non-Dutch migrants using the local paid
participant pool. Participants reported on their interactions for at
least 30 days with two daily measures (capturing the morning and
afternoon). With this design, we aimed at getting 50-60 measurements per
participant (\textit{M} = 53.26, \textit{SD} = 16.72, \textit{total N} =
1,225). This is a common number of measurements found in experience
sampling studies and offers sufficient power to model processes within
and between participants \citep[e.g.,][]{AanhetRot2012}. Participants
were compensated for their participation with up to 34 Euros -- each two
Euros for pre- and post-questionnaire and 50 Eurocents for every
experience sampling measurement. The sample consisted of relatively
young, educated, and western migrants from the global north (\(M_{age}\)
= 24.35, \(SD_{age}\) = 4.73, 19 women, 15 students). The sample
accurately describes the largest groups of migrants in the region
\citep[][]{GemeenteGroningen2015}.

@@

We recruited 71 international medical students using contacts within the
University Medical School. We specifically targeted non-Dutch students,
who had recently arrived in the Netherlands. Participants reported on
their interactions for at least 30 days with two daily measures
(capturing the morning and afternoon). With this design, we aimed at
getting 50-60 measurements per participant (\textit{M} = 57.85,
\textit{SD} = 20.68, \textit{total N} = 4,107). As with the previous
studies, this offered sufficient power to model processes within
participants. Participants were compensated in the same manner as during
Study 1. The sample consisted of relatively young migrants (\(M_{age}\)
= 22.68, \(SD_{age}\) = 3.10, 59 women). The sample fairly accurately
describes the local population of young international medical
professionals.

@@

To test whether the need mechanism was affected by whether the
interaction was accidental or planned we ran an exploratory moderation
analysis using the participants' ratings of how much they perceived the
interaction as `accidental'. It should be noted that, in order to keep
the ESM surveys short, we asked our participants to focus on most the
important interaction (i.e.,
``\textit{The following questions will be about the interaction \underline{you consider most significant}.}'';
emphasis as in original). This was also reflected in a relatively low
mean and consistent right skew of the `accidental' item. Nonetheless,
there remained a substantial variance in the item and we continued with
the moderation analysis. We found that both in the overall test as in
all studies individually key need fulfillment remained a strong
predictor of outgroup attitudes, even when accounting for differences in
whether the interaction was accidental (rather than planned). Moreover,
in none of the analysis did we see a moderation effect of interaction
intent nor did we find a main effect of interaction intent. There is,
thus, consistent evidence that the need fulfillment mechanism was not
meaningfully different for more accidental interactions.

@@

Given the well-established criticism that more positive outgroup
attitudes might not always be beneficial for minority group members
\citep[e.g.,][]{reimer2023}, we conducted an additional exploratory
analysis assessing the effect of need fulfilling outgroup interactions
on reported well-being. Experienced well-being is a common indicator of
health and an important life quality measurement in itself, especially
for migrants or other minority groups \citep[e.g.,][]{bhugra2011}. We
find that the results with well-being as the outcome variable mirror our
main results in effect size and interpretation. The results are
consistent across and within studies and, thus, add weight to the
importance of considering the need fulfillment experiences when it comes
to outgroup interactions.

>>>>

After receiving ethical approval from the University Masked for Peer
Review, we recruited 23 non-Dutch migrants using the local paid
participant pool. Participants reported on their interactions for at
least 30 days with two daily measures (capturing the morning and
afternoon). With this design, we aimed at getting 50-60 measurements per
participant (\textit{M} = 53.26, \textit{SD} = 16.72, \textit{total N} =
1,225). This is a common number of measurements found in experience
sampling studies and offers sufficient power to model processes within
and between participants \citep[e.g.,][]{AanhetRot2012}. Participants
were compensated for their participation with up to 34 Euros -- each two
Euros for pre- and post-questionnaire and 50 Eurocents for every
experience sampling measurement. The sample consisted of relatively
young, educated, and western migrants from the global north (\(M_{age}\)
= 24.35, \(SD_{age}\) = 4.73, 19 women, 15 students). The sample
accurately describes the largest groups of migrants in the region
\citep[see][for a recent report on the largest migrant groups and see Supplemental Material A for a full overview of the demographic composition, including country of origin]{GemeenteGroningen2015}.

@@

We recruited 71 international medical students using contacts within the
University Medical School. We specifically targeted non-Dutch students,
who had recently arrived in the Netherlands. Participants reported on
their interactions for at least 30 days with two daily measures
(capturing the morning and afternoon). With this design, we aimed at
getting 50-60 measurements per participant (\textit{M} = 57.85,
\textit{SD} = 20.68, \textit{total N} = 4,107). As with the previous
studies, this offered sufficient power to model processes within
participants. Participants were compensated in the same manner as during
Study 1. The sample consisted of relatively young migrants (\(M_{age}\)
= 22.68, \(SD_{age}\) = 3.10, 59 women). The sample fairly accurately
describes the local population of young international medical
professionals (see Supplemental Materials A for additional demographic information).

@@

To test whether the need mechanism was affected by whether the
interaction was accidental or planned we ran an exploratory moderation
analysis using the participants' ratings of how much they perceived the
interaction as `accidental'. It should be noted that, in order to keep
the ESM surveys short, we asked our participants to focus on most the
important interaction (i.e.,
``\textit{The following questions will be about the interaction \underline{you consider most significant}.}'';
emphasis as in original). This was also reflected in a relatively low
mean and consistent right skew of the `accidental' item. Nonetheless,
there remained a substantial variance in the item and we continued with
the moderation analysis. We found that both in the overall test and in
all studies individually key need fulfillment remained a strong
predictor of outgroup attitudes, even when accounting for differences in
whether the interaction was accidental (rather than planned). Moreover,
in none of the analyses did we see a moderation effect of interaction
intent nor did we find a main effect of interaction intent. There is,
thus, consistent evidence that the need fulfillment mechanism was not
meaningfully different for more accidental interactions.

@@

Given the well-established criticism that more positive outgroup
attitudes might not always be beneficial for minority group members
\citep[e.g.,][]{reimer2023}, we conducted an additional exploratory
analysis assessing the effect of need-fulfilling outgroup interactions
on reported well-being. Experienced well-being is a common indicator of
health and an important life quality measurement in itself, especially
for migrants or other minority groups \citep[e.g.,][]{bhugra2011}. We
find that the results with well-being as the outcome variable mirror our
main results in effect size and interpretation. The results are
consistent across and within studies and, thus, add weight to the
importance of considering the need fulfillment experiences when it comes
to outgroup interactions.

### 2023-05-31 manuscript.tex
typos

% Abstract
\abstract{
One challenge of modern intergroup contact research has been the question of when and why an interaction is perceived as positive and leads to better intergroup relations. We propose to consider the fulfillment of situationally relevant needs during everyday intergroup contact. We conducted three extensive longitudinal studies with recent migrants, to capture their interactions with the majority outgroup (total \textit{N} of measurements = 10,297; total \textit{N} participants = 207). The novel need fulfillment mechanism is consistently a strong predictor of perceived interaction quality as well as positive outgroup attitudes following intergroup interactions. The situational needs model is robust to alternative models and works at least as well as Allport's contact conditions. As one of the first studies to test intergroup contact theory using extensive longitudinal data, we offer insight into the mechanisms of positive intergroup contact during real-life interactions and find situational motivations to be a key building block of understanding and addressing positive intergroup interactions.

@@

In terms of methodological considerations, it is important to note we tested most of our hypotheses using multilevel regression models, where measurement occasions (level 1) were nested within participants (level 2). This approach is tolerant to missing data and uneven case numbers within participants. Furthermore, we use a hierarchical modeling approach and report the final model in-text \citep[][; for the full modeling process see Supplementary Material A]{snijders2012}. Secondly, statistical power estimations for intensive longitudinal- and multilevel models are notoriously difficult due to the complex covariance structures. However, our participant- and measurement numbers are among the largest sample sizes found within the intensive longitudinal literature \citep[e.g.,][]{AanhetRot2012}. Additionally, power simulations after the first study showed that our data was sufficiently powered for even small effect sizes (see Supplemental Material A). The sample sizes were only increased to allow for between-subjects comparisons and future, more complicated, time-series modeling.

@@

Finally, our conceptualization of core situational needs has been focused on the most essential test of a motivational mechanism. This comes at the expense of specificity in the situational motives (i.e., we have not explored whether different individual motives have stronger effects on interaction quality and outgroup attitudes). Such an investigation would be possible with the adaptive measurement we used (e.g., by looking at the differential effects of the clustered motives) but would not have been relevant to our theory–focused research question. Nonetheless, future research may explore which exact motives are most important in different intergroup contexts.

>>>>

% Abstract
\abstract{
One challenge of modern intergroup contact research has been the question of when and why an interaction is perceived as positive and leads to better intergroup relations. We propose to consider the perceived fulfillment of the situationally most relevant need during everyday intergroup contact. We conducted three extensive longitudinal studies with recent migrants, to capture their interactions with the majority outgroup (total \textit{N} of measurements = 10,297; total \textit{N} participants = 207). The novel core need fulfillment mechanism is consistently a strong predictor of perceived interaction quality as well as positive outgroup attitudes following intergroup interactions. The model is specific to outgroup contact, robust to various need types, and works at least as well as Allport's contact conditions. As one of the first studies to test intergroup contact theory using intensive longitudinal data, we offer insight into the mechanisms of positive intergroup contact during real-life interactions and find situational motivations to be a key building block of understanding and addressing positive intergroup interactions.

@@

In terms of methodological considerations, it is important to note we tested most of our hypotheses using multilevel regression models, where measurement occasions (level 1) were nested within participants (level 2). This approach is tolerant to missing data and uneven case numbers within participants. Furthermore, we use a hierarchical modeling approach and report the final model in-text \citep[][; for the full modeling process see Supplementary Material A]{snijders2012}. Secondly, statistical power estimations for intensive longitudinal- and multilevel models are notoriously difficult due to the complex covariance structures. However, our participant- and measurement numbers are among the largest sample sizes found within the intensive longitudinal literature \citep[e.g.,][]{AanhetRot2012}. Additionally, power simulations after the first study showed that our data was sufficiently powered for even small effect sizes (see Supplemental Material B). In particular, we found that even our smallest effects of interest would be detectable with 22 participants and 24 measurements per person (assuming the effect sizes of Study 1 and focusing on a power of .8 with a .05 alpha level; see Supplemental Material B for the full analyses). We only increased the participant sample sizes in Studies 2 and 3 to allow for between-participant effects across studies and more complicated trajectory analyses, which are not necessary for the hypotheses tested here.

@@

Finally, our conceptualization of core situational needs has been focused on the most essential test of a motivational mechanism. This comes at the expense of specificity in the situational motives (i.e., we have not explored whether different individual motives have stronger effects on interaction quality and outgroup attitudes). Such an investigation would be possible with the adaptive measurement we used (e.g., by looking at the differential effects of the clustered motives) but would not have been relevant to our theory–focused research question. Identifying cases where a specific minority faces a common need frustration could be instrumental in addressing systemic challenges. Future research may, thus, explore which situations activate or threaten specific motives \citep[e.g.,][]{gollwitzer1985e, leander2020} and which exact motives are most important in different intergroup contexts.
